Planetary perturbers: Flaring star-planet interactions in Kepler and TESS

### 2023-08-29 src/tex/ms.tex
[showyourwork] overleaf sync

% Begin!
\begin{document}

@@

\subsection{Orbital periods and transit mid-times}
\label{sec:data:orbitalperiod}
We adopt the orbital periods from the PSCP, which were either obtained from Kepler or TESS transits, or from radial velocity measurements. For the transiting systems, we use the transit mid-times given in the PSCP to set orbital phase zero. If possible, we use the transit mid-times determined in each mission separately to reduce the uncertainties on orbital phase. For the non-transiting system, phase zero is set arbitrarily.
\subsection{Semi-major axes and eccentricities}
\label{sec:data:a}
A critical parameter for the possibility of magnetic SPI and its power is the distance between the star and the planet, not the semi-major axis itself. This varies if the orbit is eccentric. We therefore adopt the mean semi-major axes from the literature, but use a custom estimate for the uncertainty that includes the eccentricity:

@@

\subsection{Planetary radii}
\label{sec:data:planetradii}
For the transiting planets, we adopt the literature values for planet radius $R_p$, and uncertainties, from the PSCP. For the non-transiting radial velocity detected planets we use  $M_p\sin i$ to calculate a lower limit for the radius using the empirical relations derived in \cite{chen2017probabilistic} using their open source \texttt{forecaster} tool, upgraded to \texttt{astro-forecaster}\footnote{https://pypi.org/project/astro-forecaster/} by Ben Cassesse. We note that for GJ 674, the mass estimates in \cite{bonfils2007harps} and \cite{boisse2011disentangling} do not quote uncertainties, but differ by $0.3M_\oplus$, so we assumed that value as the uncertainty on $M_p\sin i$.

@@

\begin{eqnarray}
    B &= 199\,\text{G} \cdot R\rm o^{-1.26\pm 0.1} \;(\text{if}\; Ro > 0.13) \label{Eq3}\\
    B &= 2050\,\text{G} \cdot R\rm o^{-0.11\pm 0.03} \;(\text{if}\; Ro < 0.13) \label{Eq4}
\end{eqnarray}

The convective turnover time $\tau$ in  $R \rm o=P_{\rm rot}/\tau$ is derived using bolometric luminosity~(Section~\ref{sec:data:lum}), following~\citet{reiners2014generalized, reiners2022magnetism}:

\begin{equation}
    \tau = 12.3\, \rm d \cdot (L_{\rm bol}/L_\odot)^{-1/2}
\end{equation}

We compared our derived values for $R$o and $B$ with existing estimates of coronal emission. We show $L_X/L_{\rm bol}$ obtained from~\citet{foster2022exoplanet} as a function of $Ro$ and $B$ in Fig.~\ref{fig:lx}. The X-ray emission relation to Rossby number in our sample follows that of stars that are not known to host close-in planets~\citep{wright2011stellaractivityrotation}, spanning both the saturated and unsaturated regime. The X-ray emission and our estimated magnetic field closely follow each other, also consistent with stars not selected for close-in planets~\citep{reiners2022magnetism}. There are two exceptions: GJ 3082 appears underluminous in X-ray compared to its Rossby number and expected magnetic field strength. AU Mic is slightly overluminous, falling somewhat above the saturated level of $L_X/L_{\rm bol}$. However, the star still follows the $B-L_X/L_{\rm bol}$ relation.

@@

\begin{equation}
    P_{\rm spi,sb} \sim B_p^{2/3} R_p^2 B_*^{4/3} R_*^4 v_{rel} a^{-4} .
\end{equation}

$B_p$ and $B_*$ are the planetary and stellar field strengths at their poles, respectively; $v_{rel}$ is the relative velocity between the stellar rotation and the planet's orbit at the semi-major axis $a$; and $R_*$ and $R_p$ are the stellar and planetary radii, respectively. Since the planetary fields are unknown, we ignore them in the calculation. We also do not know the stellar field strength at the pole for the vast majority of the stars, so we adopt the surface field strength~(Section~\ref{sec:methods:bfield}) as a proxy.

Second, we also consider the case of an unmagnetized planet with $B_p=0$, in which case Eq.~45 in~\cite{lanza2012starplanet} reduces to

\begin{equation}
    P_{\rm spi,sb0} \sim R_p^2 R_*^4 B_*^2 v_{rel}    a^{-4} 
\end{equation}

@@

\begin{equation}
    P_{\rm spi,aw} \sim B_p^{2/3} R_{p}^2  R_*^2  B_*^{1/3}  v_{rel}^2 a^{-2} \rho_*^{1/2} \sin^2 \theta
\end{equation}

@@

\begin{equation}
    P_{\rm spi,aw0} \sim R_p^2 R_*^4 B_*  v_{rel}^2 a^{-4} \rho_*^{1/2} \sin^2 \theta
\end{equation}

Lacking $B_p$, and various additional factors that appear in $P_{\rm spi,sb},P_{\rm spi,sb0},P_{\rm spi,aw}$, and  $P_{\rm spi,aw0}$, such as the angle between the relative velocity and magnetic field of the stellar wind~\citep{kavanagh2022radio}, or the force-free parameter~\citep{lanza2012starplanet}, we restrict ourselves to differential estimates, and normalize all $P_{\rm spi}$ to AU Mic.

@@

\begin{equation}
    \tau = P^2 / \sigma_P,
\end{equation}
where $P$ is the orbital period, and $\sigma_P$ is its uncertainty.
If the orbital period is well-known, the resulting ratio $\Delta T/\tau$ should be $\ll 1$. The bottom panel in Figure~\ref{fig:coherence_hist} shows a histogram of $\Delta T/\tau$ in our sample. The orbital period is known sufficiently well for our analysis in most cases with ratios $<0.02$, that is, the orbital phase of the last flare is at most $2\%$ off, compared to the first. The three cases where the ratio is highest with $0.05$, $0.12$, $0.24$ deviation are TAP 26, GJ 3082, and GJ 674, respectively. We keep these systems in our sample, but caution that the measured absence of flaring SPI might be partly due to the uncertain orbital period of the planet in these cases.

@@

\begin{table*}
\footnotesize
\movetableright=-20mm
\script{paper_main_table.py}
\caption{Flaring star-planet interaction. $Ro$, $B$, and $v_{rel}$, are derived from literature values (Table~\ref{tab:maintable_lit}). $P_{\rm spi,xx}$ stands for the power of stretch-and-break ($sb$) and Alfv\'en wing ($aw$) interaction mechanisms, assuming the planet has a magnetic field strength of $1\,$G. $P_{\rm spi,xx0}$ is the same, but assuming an unmagnetized planet. All powers are normalized to AU Mic. The $p$-value of the Anderson-Darling test is lower when the system shows a stronger clustering of flares in orbital phase.}
\input{output/table_der_vals.tex}
    \label{tab:maintable_der}

@@

\begin{table}
\script{paper_aumic_table.py}
\caption{Powers of star-planet interaction calculated for AU Mic in the four different scenarios introduced in Section~\ref{sec:methods:pspi}.}
\input{output/aumic_spi.tex}
    \label{tab:aumic}
\end{table}

@@

%The only single star system excluded is GJ 1061. The star is lacking a stellar rotation period estimate, which we need to calculate the magnetic field, and $v_{rel}$. \citet{dreizler2020reddots} estimate that it is a fairly slow rotator with a period between 50 and 200 days, so we do not expect it to have a strong magnetic field anyway. TURNS OUT IT DOES NOT HAVE ENOUGH FLARES FOR THE ANALYSIS

@@

Magnetic and tidal SPI are difficult to disentangle by their global effects on stellar activity indicators~(Section~\ref{sec:intro:global}). Locally, however, we can discriminate between the two by their relevant periods, that is, $P_{\rm orb}$ and $P_{\rm orb /2}$. In Section~\ref{sec:discussion:tidal}, we repeat the analysis in this paper with $P_{\rm orb}/2$, and find yet more tentative trends in line with scaling laws for tidal interaction. In a system-by-system comparison, we find consistency with the expectation that slowly rotating, likely old, systems can only show local tidal interaction, while rapidly rotating systems with strong magnetic fields are dominated by magnetic forces.
\subsection{Alfv\'en surface}
\label{sec:discussion:as}

@@

\subsection{Tidal interaction}
\label{sec:discussion:tidal}
\begin{figure*}[t]
    \script{paper_tidal_interaction.py}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/PAPER_tidal_AD_test.png}
    \caption{Deviation from random flare times in phase with $P_{\rm orb}/2$, i.e., in phase with the two tidal bulges raised on the star, compared to different models of tidal interaction~(Section~\ref{sec:discussion:tidal}). \textbf{Cross symbols} mark planets with $M_p \sin i$ measurements. \textbf{Round symbols} mark planets with measured  or estimated $M_p$ (see Section~\ref{sec:discussion:mpms}). 
           \textbf{Left panel}: Tidal torque in the system for stars with a convective envelope. Blue color indicates that the star rotates slower than the planet orbits, so that angular momentum is transferred from the orbit to the spin of the star. Grey color is the reverse. \textbf{Middle panel}: The tidal dissipation timescale. \textbf{Right panel}:  Relative gravitational perturbation in the star.}
        \label{fig:tidal}
    \end{centering}
\end{figure*}

@@

\subsubsection{Planetary and stellar masses}
\label{sec:discussion:mpms}
We collect the planetary and stellar masses from the literature~(Table~\ref{tab:tidal}). For missing planetary masses, the radii were known, so we used the mass-radius relations implemented in \texttt{astro-forecaster}~\citep[Ben Cassesse's implementation of \texttt{forecaster},][]{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute $K_s$ magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. For Kepler-42 b, we used  $M_p=0.1-2.06M_\oplus$ that covers pure rock to pure iron compositions in \citet{muirhead2012characterizing}, and adopted the logarithmic mean for our estimate. Kepler-1558 c has a similar radius to Kepler-42 b, i.e., most likely a bare core planet, so we used density ranging from Earth bulk density to pure rock to estimate its mass. We also adopt the upper limit of $5M_J$ for HIP 67522 from~\citet{rizzuto2020tess}, but use the lower error from \citet{chen2017probabilistic}. The assumptions we made here are conservative, and the estimates based on \citet{chen2017probabilistic} use a broad sample of planets, so we consider the planetary mass estimates with their respective uncertainties robust. We note, however, that the mass estimates for RV-detected planets are $M_p \sin i$, so that their expected interaction may be stronger than indicated in Fig.~\ref{fig:tidal}.

>>>>

% Begin!
\usepackage{makecell}
\begin{document}

@@

\subsection{Orbital periods and transit mid-times}
\label{sec:data:orbitalperiod}
We adopt the orbital periods from the PSCP, which were either obtained from Kepler or TESS transits, or from radial velocity measurements. For the transiting systems, we use the transit mid-times given in the PSCP to set orbital phase zero. If possible, we use the transit mid-times determined in each mission separately to reduce the uncertainties on orbital phase. For the non-transiting system, phase zero is set arbitrarily.

\subsection{Semi-major axes and eccentricities}
\label{sec:data:a}
A critical parameter for the possibility of magnetic SPI and its power is the distance between the star and the planet, not the semi-major axis itself. This varies if the orbit is eccentric. We therefore adopt the mean semi-major axes from the literature, but use a custom estimate for the uncertainty that includes the eccentricity:

@@

\subsection{Planetary radii}
\label{sec:data:planetradii}
For the transiting planets, we adopt the literature values for planet radius $R_{\rm p}$, and uncertainties, from the PSCP. For the non-transiting radial velocity detected planets we use  $M_{\rm p}\sin i$ to calculate a lower limit for the radius using the empirical relations derived in \cite{chen2017probabilistic} using their open source \texttt{forecaster} tool, upgraded to \texttt{astro-forecaster}\footnote{https://pypi.org/project/astro-forecaster/} by Ben Cassesse. We note that for GJ 674, the mass estimates in \cite{bonfils2007harps} and \cite{boisse2011disentangling} do not quote uncertainties, but differ by $0.3M_\oplus$, so we assumed that value as the uncertainty on $M_{\rm p}\sin i$.

@@

\begin{eqnarray}
    B &= 199\,\text{G} \cdot R\mathrm{o}^{-1.26\pm 0.1} \;(\text{if}\; R\mathrm{o} > 0.13) \label{Eq3}\\
    B &= 2050\,\text{G} \cdot R\mathrm{o}^{-0.11\pm 0.03} \;(\text{if}\; R\mathrm{o} < 0.13) \label{Eq4}
\end{eqnarray}

The convective turnover time $\tau$ in  $R \mathrm{o}=P_{\rm rot}/\tau$ is derived using bolometric luminosity~(Section~\ref{sec:data:lum}), following~\citet{reiners2014generalized, reiners2022magnetism}:

\begin{equation}
    \tau = 12.3\, \mathrm{d} \cdot (L_{\rm bol}/L_\odot)^{-1/2}
\end{equation}

We compared our derived values for $R$o and $B$ with existing estimates of coronal emission. We show $L_X/L_{\rm bol}$ obtained from~\citet{foster2022exoplanet} as a function of $R$o and $B$ in Fig.~\ref{fig:lx}. The X-ray emission relation to Rossby number in our sample follows that of stars that are not known to host close-in planets~\citep{wright2011stellaractivityrotation}, spanning both the saturated and unsaturated regime. The X-ray emission and our estimated magnetic field closely follow each other, also consistent with stars not selected for close-in planets~\citep{reiners2022magnetism}. There are two exceptions: GJ 3082 appears underluminous in X-ray compared to its Rossby number and expected magnetic field strength. AU Mic is slightly overluminous, falling somewhat above the saturated level of $L_X/L_{\rm bol}$. However, the star still follows the $B-L_X/L_{\rm bol}$ relation.

@@

\begin{equation}
    P_{\rm spi,sb} \sim B_{\rm p}^{2/3} R_{\rm p}^2 B_*^{4/3} R_*^4 v_{\rm rel} a^{-4} .
\end{equation}

$B_{\rm p}$ and $B_*$ are the planetary and stellar field strengths at their poles, respectively; $v_{\rm rel}$ is the relative velocity between the stellar rotation and the planet's orbit at the semi-major axis $a$; and $R_*$ and $R_{\rm p}$ are the stellar and planetary radii, respectively. Since the planetary fields are unknown, we ignore them in the calculation. We also do not know the stellar field strength at the pole for the vast majority of the stars, so we adopt the surface field strength~(Section~\ref{sec:methods:bfield}) as a proxy.

Second, we also consider the case of an unmagnetized planet with $B_{\rm p}=0$, in which case Eq.~45 in~\cite{lanza2012starplanet} reduces to

\begin{equation}
    P_{\rm spi,sb0} \sim R_{\rm p}^2 R_*^4 B_*^2 v_{\rm rel}    a^{-4} 
\end{equation}

@@

\begin{equation}
    P_{\rm spi,aw} \sim B_{\rm p}^{2/3} R_{p}^2  R_*^2  B_*^{1/3}  v_{\rm rel}^2 a^{-2} \rho_*^{1/2} \sin^2 \theta
\end{equation}

@@

\begin{equation}
    P_{\rm spi,aw0} \sim R_{\rm p}^2 R_*^4 B_*  v_{\rm rel}^2 a^{-4} \rho_*^{1/2} \sin^2 \theta
\end{equation}

Lacking $B_{\rm p}$, and various additional factors that appear in $P_{\rm spi,sb},P_{\rm spi,sb0},P_{\rm spi,aw}$, and  $P_{\rm spi,aw0}$, such as the angle between the relative velocity and magnetic field of the stellar wind~\citep{kavanagh2022radio}, or the force-free parameter~\citep{lanza2012starplanet}, we restrict ourselves to differential estimates, and normalize all $P_{\rm spi}$ to AU Mic.

@@

\begin{equation}
    \tau = P^2 / \sigma_P,
\end{equation}
where $P$ is the orbital period, and $\sigma_P$ is its uncertainty.
If the orbital period is well-known, the resulting ratio $\Delta T/\tau$ should be $\ll 1$. \textcolor{red}{Here, we are not interested in the absolute orbital phase, but only in the relative one. Therefore, we can ignore that the uncertainty on phase zero at transit midtime increases with the time passed since the transit midtime was measured. For non-transiting systems, this time can amount to several years, and the midtime is unconstrained by the time of the observation with Kepler or TESS. For transiting systems, the midtime is usually very well known. So in the following, we set phsae zero to an arbitrary time in non-transiting systems, and to the reported transit midtime for the rest.} The bottom panel in Figure~\ref{fig:coherence_hist} shows a histogram of $\Delta T/\tau$ in our sample. The orbital period is known sufficiently well for our analysis in most cases with ratios $<0.02$, that is, the orbital phase of the last flare is at most $2\%$ off, compared to the first. The three cases where the ratio is highest with $0.05$, $0.12$, $0.24$ deviation are TAP 26, GJ 3082, and GJ 674, respectively. We keep these systems in our sample, but caution that the measured absence of flaring SPI might be partly due to the uncertain orbital period of the planet in these cases.

@@

\begin{table*}
\footnotesize
\movetableright=-20mm
\script{paper_main_table.py}
\caption{Flaring star-planet interaction. $Ro$, $B$, and $v_{\rm rel}$, are derived from literature values (Table~\ref{tab:maintable_lit}). $P_{\rm spi,xx}$ stands for the power of stretch-and-break ($sb$) and Alfv\'en wing ($aw$) interaction mechanisms, assuming the planet has a magnetic field strength of $1\,$G. $P_{\rm spi,xx0}$ is the same, but assuming an unmagnetized planet. All powers are normalized to AU Mic. The $p$-value of the Anderson-Darling test is lower when the system shows a stronger clustering of flares in orbital phase.}
\input{output/table_der_vals.tex}
    \label{tab:maintable_der}

@@

\begin{table}
\movetableright=-10mm
\script{paper_aumic_table.py}
\caption{\textcolor{red}{Powers of star-planet interaction calculated for AU Mic in the four different scenarios introduced in Section~\ref{sec:methods:pspi}.}}
\input{output/aumic_spi.tex}
    \label{tab:aumic}
\end{table}

@@

%The only single star system excluded is GJ 1061. The star is lacking a stellar rotation period estimate, which we need to calculate the magnetic field, and $v_{\rm rel}$. \citet{dreizler2020reddots} estimate that it is a fairly slow rotator with a period between 50 and 200 days, so we do not expect it to have a strong magnetic field anyway. TURNS OUT IT DOES NOT HAVE ENOUGH FLARES FOR THE ANALYSIS

@@

Magnetic and tidal SPI are difficult to disentangle by their global effects on stellar activity indicators~(Section~\ref{sec:intro:global}). Locally, however, we can discriminate between the two by their relevant periods, that is, $P_{\rm orb}$ and $P_{\rm orb} /2$. In Section~\ref{sec:discussion:tidal}, we repeat the analysis in this paper with $P_{\rm orb}/2$, and find yet more tentative trends in line with scaling laws for tidal interaction. In a system-by-system comparison, we find consistency with the expectation that slowly rotating, likely old, systems can only show local tidal interaction, while rapidly rotating systems with strong magnetic fields are dominated by magnetic forces.
\subsection{Alfv\'en surface}
\label{sec:discussion:as}

@@

\subsection{Tidal interaction}
\label{sec:discussion:tidal}
\begin{figure*}[t]
    \script{paper_tidal_interaction.py}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/PAPER_tidal_AD_test.png}
    \caption{Deviation from random flare times in phase with $P_{\rm orb}/2$, i.e., in phase with the two tidal bulges raised on the star, compared to different models of tidal interaction~(Section~\ref{sec:discussion:tidal}). \textbf{Cross symbols} mark planets with $M_{\rm p} \sin i$ measurements. \textbf{Round symbols} mark planets with measured  or estimated $M_{\rm p}$ (see Section~\ref{sec:discussion:mpms}). 
           \textbf{Left panel}: Tidal torque in the system for stars with a convective envelope. Blue color indicates that the star rotates slower than the planet orbits, so that angular momentum is transferred from the orbit to the spin of the star. Grey color is the reverse. \textbf{Middle panel}: The tidal dissipation timescale. \textbf{Right panel}:  Relative gravitational perturbation in the star.}
        \label{fig:tidal}
    \end{centering}
\end{figure*}

@@

\subsubsection{Planetary and stellar masses}
\label{sec:discussion:mpms}
We collect the planetary and stellar masses from the literature~(Table~\ref{tab:tidal}). For missing planetary masses, the radii were known, so we used the mass-radius relations implemented in \texttt{astro-forecaster}~\citep[Ben Cassesse's implementation of \texttt{forecaster},][]{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute $K_{\rm s}$ magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. For Kepler-42 b, we used  $M_{\rm p}=0.1-2.06M_\oplus$ that covers pure rock to pure iron compositions in \citet{muirhead2012characterizing}, and adopted the logarithmic mean for our estimate. Kepler-1558 c has a similar radius to Kepler-42 b, i.e., most likely a bare core planet, so we used density ranging from Earth bulk density to pure rock to estimate its mass. We also adopt the upper limit of $5M_{\rm J}$ for HIP 67522 from~\citet{rizzuto2020tess}, but use the lower error from \citet{chen2017probabilistic}. The assumptions we made here are conservative, and the estimates based on \citet{chen2017probabilistic} use a broad sample of planets, so we consider the planetary mass estimates with their respective uncertainties robust. We note, however, that the mass estimates for RV-detected planets are $M_{\rm p} \sin i$, so that their expected interaction may be stronger than indicated in Fig.~\ref{fig:tidal}.
